Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-06-20 conclusions.tex
Mike Norman's edits.

With the continual rapid development of computer hardware, it makes
sense to not only review \enzo's current capabilities, but to look
toward its future development in view of predicted technological
trends. These trends in supercomputing hardware suggest that
substantial modifications to \enzo's core infrastructure, and very
possibly some of the core algorithms, will be required. More
specifically, the progression involves the usage of specialized
large-core-count, vectorized computing units such as graphics
processing units or chips like the Intel Xeon Phi, as well as
precipitously decreasing amounts of RAM per computing core.  The
former trend means that the amount of processing power per compute
node will continue to increase, likely much faster than the bandwidth
between nodes, and will require tremendous reduction in inter-node
(and possibly inter-CPU) communication in order to maintain code
scalability.  Also, much of the current code will need to be rewritten
to take advantage of the vector nature of these CPUs, making
assumptions that are quite unlike those made in much of the current
codebase.  The latter trend means that duplication of data -- for
example, the grid hierarchy -- must be effectively eliminated to save
memory, and all inter-core and inter-node communication must be
carefully thought through to minimize the amount of data moved.  An
additional challenge as one goes to core counts in the tens to
hundreds of millions (or more) is that the reliability of individual
computing elements will become much more of an issue, requiring
robustness to hardware failure to be built into the code at some
level.  Furthermore, we are nearing the physical limits of transistor
speed and latency~\citep{feynman1999feynman}, meaning that simple
hardware improvements will not make these challenges disappear, and
careful thought (and the rewriting of a great deal of code) must take
place!  These challenges are not unique to the \enzo\ code, and in
fact are faced by effectively all applications that wish to take
advantage of new computational architectures. We therefore anticipated
that \enzo\ (or a code that has the capabilities of \enzo, from a
user's point of view) will continue to be usable at the
largest scales on such machines.

>>>>

With the continual rapid development of computer hardware, it makes
sense to not only review \enzo's current capabilities, but to look
toward its future development in view of predicted technological
trends. These trends in supercomputing hardware suggest that
substantial modifications to \enzo's core infrastructure, and very
possibly some of the core algorithms, will be required. More
specifically, the progression involves the usage of specialized
large-core-count, vectorized computing units such as graphics
processing units or chips like the Intel Xeon Phi, as well as
precipitously decreasing amounts of RAM per computing core.  The
former trend means that the amount of processing power per compute
node will continue to increase, likely much faster than the bandwidth
between nodes, and will require tremendous reduction in inter-node
(and possibly inter-CPU) communication in order to maintain code
scalability.  Also, much of the current code will need to be rewritten
to take advantage of the vector nature of these CPUs, making
assumptions that are quite unlike those made in much of the current
codebase.  The latter trend means that duplication of data -- for
example, the grid hierarchy -- must be effectively eliminated to save
memory, and all inter-core and inter-node communication must be
carefully thought through to minimize the amount of data moved.  An
additional challenge as one goes to core counts in the tens to
hundreds of millions (or more) is that the reliability of individual
computing elements will become much more of an issue, requiring
robustness to hardware failure to be built into the code at some
level.  Furthermore, we are nearing the physical limits of transistor
speed and latency~\citep{feynman1999feynman}, meaning that simple
hardware improvements will not make these challenges disappear, and
careful thought (and the rewriting of a great deal of code) must take
place!  These challenges are not unique to the \enzo\ code, and in
fact are faced by effectively all applications that wish to take
advantage of new computational architectures. We therefore anticipate
that \enzo\ (or a code that has the capabilities of \enzo, from a
user's point of view) will continue to be usable at the
largest scales on such machines.

### 2013-06-20 introduction.tex
Mike Norman's edits.

Due to the high spatial and temporal dynamical ranges involved,
astrophysics and cosmological phenomena present a taxing challenge for
simulators. To tackle such situations, a number of numerical
techniques have been developed that can be broadly split into the
gridless, Lagrangian methods and grid-based, Eulerian schemes. The
most commonly used is an example of the first type known as Smoothed
Particle Hydrodynamics \citep[SPH;][]{Lucy77, SPH}. It has achieved
much success, particularly in regimes dominated by gravity. However,
its development to include an increasing number of sought-after
physical processes is still at a relatively early stage when compared
with the effort put into the latter type of Eulerian grid-based
hydrodynamic schemes \citep[e.g.,][]{laney-1998, toro-1997,
Woodward84}.

Despite this invested expertise, the Eulerian solvers in their
original form have a serious drawback: they do not provide an easy
method of adaptively increasing the spatial and temporal resolution in
small volumes of the simulation. Such flexibility is essential for
following physical processes such as gravitational instability. A
solution to this was first proposed by \citet{Berger89} in the
Computational Fluid Dynamics (CFD) community, and became known as
Structured Adaptive Mesh Refinement (SAMR). The principle is to
adaptively add and modify additional, finer meshes (``grids'') over
regions that require higher resolution. In addition, it is possible to
add other, more advanced physics including -- for the AMR
implementation in the astrophysics code presented in this paper --
comoving coordinates, self-gravity, radiative cooling, chemistry, heat
conduction, collisionless fluids, magnetohydrodynamics, radiation
transport, star formation and a range of other physical effects.
\newpage

>>>>

Due to the high spatial and temporal dynamical ranges involved,
astrophysical and cosmological phenomena present a taxing challenge for
simulators. To tackle such situations, a number of numerical
techniques have been developed that can be broadly split into the
gridless, Lagrangian methods and grid-based, Eulerian schemes. The
most commonly used is an example of the first type known as Smoothed
Particle Hydrodynamics \citep[SPH;][]{Lucy77, SPH}. It has achieved
much success, particularly in regimes dominated by gravity. However,
its development to include an increasing number of sought-after
physical processes is still at a relatively early stage when compared
with the effort put into the latter type of Eulerian grid-based
hydrodynamic schemes \citep[e.g.,][]{laney-1998, toro-1997,
Woodward84}.

Despite this invested expertise, the Eulerian solvers in their
original form have a serious drawback: they do not provide an easy
method of adaptively increasing the spatial and temporal resolution in
small volumes of the simulation. Such flexibility is essential for
following physical processes such as gravitational instability. A
solution to this problem was first proposed by \citet{Berger89} in the
Computational Fluid Dynamics (CFD) community, and became known as
Structured Adaptive Mesh Refinement (SAMR). The principle is to
adaptively add and modify additional, finer meshes (``grids'') over
regions that require higher resolution. In addition, it is possible to
add other, more advanced physics including -- for the AMR
implementation in the astrophysics code presented in this paper --
comoving coordinates, self-gravity, radiative cooling, chemistry, heat
conduction, collisionless fluids, magnetohydrodynamics, radiation
transport, star formation and a range of other physical effects.
\newpage

### 2013-06-20 ms.tex
Fixed extra comma after Enzo Collab in author list
Updated SMU affiliation line
Christine Simpson's changes.
adding stephen skory changes
Mike Norman's edits.

\title{Enzo: An Adaptive Mesh Refinement Code for Astrophysics}
\author{The Enzo Collaboration:}
\author{Greg L. Bryan\altaffilmark{1}}
\author{Michael L. Norman\altaffilmark{2,3}}
\author{Brian W. O'Shea\altaffilmark{4,5}}
\author{Tom Abel\altaffilmark{6, 20}}
\author{John H. Wise\altaffilmark{7}}
\author{Matthew J. Turk\altaffilmark{1}}
\author{Daniel R. Reynolds\altaffilmark{8}}
\author{David C. Collins\altaffilmark{9}}
\author{Peng Wang\altaffilmark{6}}
\author{Samuel W. Skillman\altaffilmark{10,11}}
\author{Britton Smith\altaffilmark{4}}
\author{Robert P. Harkness\altaffilmark{12}}
\author{James Bordner\altaffilmark{2}}
\author{Ji-hoon Kim\altaffilmark{13}}
\author{Michael Kuhlen\altaffilmark{14,15}}
\author{Hao Xu\altaffilmark{2}}
\author{Nathan Goldbaum\altaffilmark{15}}
\author{Cameron Hummels\altaffilmark{16}}
\author{Alexei G. Kritsuk\altaffilmark{2}}
\author{Elizabeth Tasker\altaffilmark{17}}
\author{Stephen Skory\altaffilmark{2}}
\author{Christine Simpson\altaffilmark{1}}
\author{Oliver Hahn\altaffilmark{18}}
\author{Jeffrey S. Oishi\altaffilmark{19}}
\author{Geoffrey C So\altaffilmark{2}}
\author{Fen Zhao\altaffilmark{20}}
\author{Renyue Cen\altaffilmark{21}}
\author{Yuan Li\altaffilmark{1}}

\altaffiltext{1}{Columbia University, Department of Astronomy, New York, NY, 10025, USA}
\altaffiltext{2}{CASS, University of California, San Diego, 9500 Gilman Drive La Jolla, CA 92093-0424, USA}
\altaffiltext{3}{SDSC, University of California, San Diego, 9500 Gilman Drive La Jolla, CA 92093-0505, USA}
\altaffiltext{4}{Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\altaffiltext{5}{Lyman Briggs College and Institute for Cyber-Enabled Research, Michigan State University, East Lansing, MI 48824, USA}
\altaffiltext{6}{Kavli Institute for Particle Astrophysics and Cosmology, Stanford University, Menlo Park, CA 94025, USA}
\altaffiltext{7}{Center for Relativistic Astrophysics, School of Physics, Georgia Institute of Technology, 837 State Street, Atlanta, GA, USA}
\altaffiltext{8}{Southern Methodist University, 6425 Boaz Ln, Dallas, TX 75205, USA}
\altaffiltext{9}{Theoretical Division, Los Alamos National Laboratory, Los Alamos, NM 87544}
\altaffiltext{10}{Center for Astrophysics and Space Astronomy, Department of Astrophysical and Planetary Science, University of Colorado, Boulder, CO 80309, USA}
\altaffiltext{11}{DOE Computational Science Graduate Fellow}
\altaffiltext{12}{NICS, Oak Ridge National Laboratory, 1 Bethel Valley Rd, Oak Ridge, TN 37831, USA}
\altaffiltext{13}{Department of Astronomy and Astrophysics, University of California, Santa Cruz, CA 95064, USA}
\altaffiltext{14}{Theoretical Astrophysics Center, University of California Berkeley, Hearst Field Annex, Berkeley, CA 94720, USA}
\altaffiltext{15}{Department of Astronomy and Astrophysics, University of California, 1156 High Street, Santa Cruz, CA 95064, USA}
\altaffiltext{16}{Department of Astronomy/Steward Observatory, University of Arizona, 933 N. Cherry Ave., Tucson, AZ, 85721, USA}
\altaffiltext{17}{Physics Department, Faculty of Science, Hokkaido University, Kita-10 Nishi 8, Kita-ku, Sapporo 060-0810, Japan}
\altaffiltext{18}{ETH Zurich Institute for Astronomy. CH-8093 Zurich. Switzerland}
\altaffiltext{19}{Department of Astrophysics, American Museum of Natural History, Central Park West at 81st St, New York, NY 10024}
\altaffiltext{20}{Department of Physics, Stanford University, Stanford, CA 94305, USA}
\altaffiltext{21}{Department of Astrophysical Sciences, Princeton University, Peyton Hall, Princeton, NJ 08544, USA}

\begin{abstract}
This paper describes the open-source code \enzo, which uses a
block-structured adaptive mesh refinement to provide high spatial and
temporal resolution for modeling astrophysical fluid flows.  The code
is Cartesian and can be run in 1, 2, and 3 dimensions, and supports a
large variety of physics, including hydrodynamics, ideal and non-ideal
magnetohydrodynamics, N-body dynamics (and, more broadly, self-gravity
of fluids and particles), primordial gas chemistry, optically-thin
radiative cooling of primordial and metal-enriched plasmas (as well as
some optically-thick cooling models), radiation
transport, cosmological expansion, and models for star formation and
feedback in a cosmological context.  In addition to explaining the
algorithms implemented, we present solutions for a wide range of test
problems, demonstrate the code's parallel scaling properties, and
discuss the \enzo\ collaboration's code development methodology.
\end{abstract}

>>>>

\title{Enzo: An Adaptive Mesh Refinement Code for Astrophysics}
\author{The Enzo Collaboration: Greg L. Bryan\altaffilmark{1}}
\author{Michael L. Norman\altaffilmark{2,3}}
\author{Brian W. O'Shea\altaffilmark{4,5}}
\author{Tom Abel\altaffilmark{6, 20}}
\author{John H. Wise\altaffilmark{7}}
\author{Matthew J. Turk\altaffilmark{1}}
\author{Daniel R. Reynolds\altaffilmark{8}}
\author{David C. Collins\altaffilmark{9}}
\author{Peng Wang\altaffilmark{6}}
\author{Samuel W. Skillman\altaffilmark{10,11}}
\author{Britton Smith\altaffilmark{4}}
\author{Robert P. Harkness\altaffilmark{12}}
\author{James Bordner\altaffilmark{2}}
\author{Ji-hoon Kim\altaffilmark{13}}
\author{Michael Kuhlen\altaffilmark{14,15}}
\author{Hao Xu\altaffilmark{2}}
\author{Nathan Goldbaum\altaffilmark{15}}
\author{Cameron Hummels\altaffilmark{16}}
\author{Alexei G. Kritsuk\altaffilmark{2}}
\author{Elizabeth Tasker\altaffilmark{17}}
\author{Stephen Skory\altaffilmark{10}}
\author{Christine M. Simpson\altaffilmark{1}}
\author{Oliver Hahn\altaffilmark{18}}
\author{Jeffrey S. Oishi\altaffilmark{19}}
\author{Geoffrey C So\altaffilmark{2}}
\author{Fen Zhao\altaffilmark{20}}
\author{Renyue Cen\altaffilmark{21}}
\author{Yuan Li\altaffilmark{1}}

\altaffiltext{1}{Columbia University, Department of Astronomy, New York, NY, 10025, USA}
\altaffiltext{2}{CASS, University of California, San Diego, 9500 Gilman Drive La Jolla, CA 92093-0424, USA}
\altaffiltext{3}{SDSC, University of California, San Diego, 9500 Gilman Drive La Jolla, CA 92093-0505, USA}
\altaffiltext{4}{Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\altaffiltext{5}{Lyman Briggs College and Institute for Cyber-Enabled Research, Michigan State University, East Lansing, MI 48824, USA}
\altaffiltext{6}{Kavli Institute for Particle Astrophysics and Cosmology, Stanford University, Menlo Park, CA 94025, USA}
\altaffiltext{7}{Center for Relativistic Astrophysics, School of Physics, Georgia Institute of Technology, 837 State Street, Atlanta, GA, USA}
\altaffiltext{8}{Department of Mathematics, Southern Methodist University, Box 750156, Dallas, TX 75205-0156, USA}
\altaffiltext{9}{Theoretical Division, Los Alamos National Laboratory, Los Alamos, NM 87544}
\altaffiltext{10}{Center for Astrophysics and Space Astronomy, Department of Astrophysical and Planetary Science, University of Colorado, Boulder, CO 80309, USA}
\altaffiltext{11}{DOE Computational Science Graduate Fellow}
\altaffiltext{12}{NICS, Oak Ridge National Laboratory, 1 Bethel Valley Rd, Oak Ridge, TN 37831, USA}
\altaffiltext{13}{Department of Astronomy and Astrophysics, University of California, Santa Cruz, CA 95064, USA}
\altaffiltext{14}{Theoretical Astrophysics Center, University of California Berkeley, Hearst Field Annex, Berkeley, CA 94720, USA}
\altaffiltext{15}{Department of Astronomy and Astrophysics, University of California, 1156 High Street, Santa Cruz, CA 95064, USA}
\altaffiltext{16}{Department of Astronomy/Steward Observatory, University of Arizona, 933 N. Cherry Ave., Tucson, AZ, 85721, USA}
\altaffiltext{17}{Physics Department, Faculty of Science, Hokkaido University, Kita-10 Nishi 8, Kita-ku, Sapporo 060-0810, Japan}
\altaffiltext{18}{ETH Zurich Institute for Astronomy. CH-8093 Zurich. Switzerland}
\altaffiltext{19}{Department of Astrophysics, American Museum of Natural History, Central Park West at 81st St, New York, NY 10024}
\altaffiltext{20}{Department of Physics, Stanford University, Stanford, CA 94305, USA}
\altaffiltext{21}{Department of Astrophysical Sciences, Princeton University, Peyton Hall, Princeton, NJ 08544, USA}

\begin{abstract}
This paper describes the open-source code \enzo, which uses 
block-structured adaptive mesh refinement to provide high spatial and
temporal resolution for modeling astrophysical fluid flows.  The code
is Cartesian and can be run in 1, 2, and 3 dimensions, and supports a
large variety of physics, including hydrodynamics, ideal and non-ideal
magnetohydrodynamics, N-body dynamics (and, more broadly, self-gravity
of fluids and particles), primordial gas chemistry, optically-thin
radiative cooling of primordial and metal-enriched plasmas (as well as
some optically-thick cooling models), radiation
transport, cosmological expansion, and models for star formation and
feedback in a cosmological context.  In addition to explaining the
algorithms implemented, we present solutions for a wide range of test
problems, demonstrate the code's parallel scaling properties, and
discuss the \enzo\ collaboration's code development methodology.
\end{abstract}

### 2013-06-20 numerical-ppm.tex
Mike Norman's edits.

In this section, we describe the four solvers that we have implemented for
solving the fluid equations.  We describe the PPM method in considerably
more detail than the other methods in part
because its implementation in Enzo has not previously been described,
but mostly because it introduces many of the ideas and methods used for
the MUSCL-Dedner and MHDCT schemes (including expansion terms,
reconstruction, Riemann solvers, and dual energy formalism);

>>>>

In this section, we describe the four solvers that we have implemented for
solving the fluid equations.  We describe the PPM method in considerably
more detail than the other methods in part
because its implementation in Enzo has not previously been described,
but mostly because it introduces many of the ideas and methods used for
the MUSCL-Dedner and MHDCT schemes (including expansion terms,
reconstruction, Riemann solvers, and dual energy formalism).

### 2013-06-20 numerical-zeus.tex
Mike Norman's edits.

The \zeus\ method uses a von Neumann-Richtmeyer artificial viscosity to smooth shock discontinuities that may appear in fluid flows and can cause a breakdown of the finite-difference equations.  The artificial viscosity term is added in the source terms as:
\begin{eqnarray}
\rho \frac{\partial\vecv}{\partial t} &=& - \grad p - \rho \grad \phi 
- \div \textbf{Q} \\
\frac{\partial e}{\partial t} &=& -p \div \vecv - \textbf{Q} : \grad \vecv, 
\end{eqnarray}
Here \textbf{Q} is the artificial viscosity stress tensor, which we take to be diagonal with on-axis terms given by $l^2 \rho (\partial v / \partial x)^2$ as proposed by von Neumann \& Richtmyer.  The length scale $l$ determines the width of shocks and is typically a few times the cell spacing.

@@

Next, we examine the transport step, which is conservative.  Once again, we dimensionally split the equations and present only the one-dimensional version.  The finite-difference equations actually solved are:
\begin{equation}
\rho_j^{n+d} = \rho_j^{n} - \frac{\Delta t}{\Delta x} (v^{n+c}_{j+1/2} \rho^{*}_{j+1/2} - v^{n+c}_{j-1/2} \rho^{*}_{j-1/2} )
\end{equation}
Here $\rho^*_j$ is the correctly upwinded value of $\rho$ evaluated at the cell-face corresponding to $v_j$, making $\rho^*_j v_j$ the mass flux at the cell boundary and guaranteeing mass conservation.   This requires interpolating each cell-centered quantity to the cell edge.  As recomended in \citet{Stone92a}, we use the second-order van Leer scheme, which uses piecewise linear functions.  These are given by equations (48) and (49) of \citet{Stone92a}.  The transport steps for the other variables are similar.  Note that we advect the specific energy and specific momenta using the mass flux, as dictated by the principle of consistent transport.  This requires appropriate averaging for the momenta in the perpendicular directions as outlined in equations (57)-(72) of \citet{Stone92a}.

>>>>

The \zeus\ method uses a von Neumann-Richtmyer artificial viscosity to smooth shock discontinuities that may appear in fluid flows and can cause a breakdown of the finite-difference equations.  The artificial viscosity term is added in the source terms as:
\begin{eqnarray}
\rho \frac{\partial\vecv}{\partial t} &=& - \grad p - \rho \grad \phi 
- \div \textbf{Q} \\
\frac{\partial e}{\partial t} &=& -p \div \vecv - \textbf{Q} : \grad \vecv, 
\end{eqnarray}
Here \textbf{Q} is the artificial viscosity stress tensor, which we take to be diagonal with on-axis terms given by $l^2 \rho (\partial v / \partial x)^2$ as proposed by von Neumann \& Richtmyer.  The length scale $l$ determines the width of shocks and is typically a few times the cell spacing.

@@

Next, we examine the transport step, which is conservative.  Once again, we dimensionally split the equations and present only the one-dimensional version.  The finite-difference equations actually solved are:
\begin{equation}
\rho_j^{n+d} = \rho_j^{n} - \frac{\Delta t}{\Delta x} (v^{n+c}_{j+1/2} \rho^{*}_{j+1/2} - v^{n+c}_{j-1/2} \rho^{*}_{j-1/2} )
\end{equation}
Here $\rho^*_j$ is the correctly upwinded value of $\rho$ evaluated at the cell-face corresponding to $v_j$, making $\rho^*_j v_j$ the mass flux at the cell boundary and guaranteeing mass conservation.   This requires interpolating each cell-centered quantity to the cell edge.  As recommended in \citet{Stone92a}, we use the second-order van Leer scheme, which uses piecewise linear functions.  These are given by equations (48) and (49) of \citet{Stone92a}.  The transport steps for the other variables are similar.  Note that we advect the specific energy and specific momenta using the mass flux, as dictated by the principle of consistent transport.  This requires appropriate averaging for the momenta in the perpendicular directions as outlined in equations (57)-(72) of \citet{Stone92a}.

### 2013-06-20 overview.tex
Mike Norman's edits.

In addition, \enzo\ can solve the mass conservation equations for a set of
chemical species and their reactions.  For any species $i$ with comoving
number density $n_i$, these equations have the form:
\begin{equation}
  \frac{\partial n_i}{\partial t} 
  + \frac{1}{a} \div (n_i \vecv) = 
  \sum_j k_{ij}(T) \, n_i n_j + \sum_j \Gamma^{\rm ph}_j n_j 
  \label{eq:species_evolution}
\end{equation}
where $k_{ij}$ are the rate coefficients for the two-body reactions
and are usually functions of only the gas species (we will
specifically note the cases where we either include three-body
reactions or have density-dependent rates).  The $\Gamma^{\rm ph}_j$
are destruction/creation rates due to photoionizations and/or
photodissociations.  Currently the species \enzo\ can follow include
H, H$^+$, He, He$^+$, He$^{++}$, and e$^-$, and through additional
options also H$^-$, H$_2$, H$_2^+$, and HD, D, and D$^+$. Lastly,
\enzo\ can also track the advection of a comoving metal density field,
which can contribute to the radiative cooling and star formation
processes.

@@

In this section we briefly describe the numerical methods which are
used to solve the equations outlined in Section~\ref{sec.equations}.
We proceed through the numerical methods in the same order as will be
used in Sections~\ref{sec.amr} through~\ref{sec.num.analysis} so that
there is a one-to-one correspondence between each of the following
overviews and the complete description provided in the later sections.
The goal in this section is to introduce the reader to the basic
principles of the methods without drowning in detail.

@@

As described in \citet{Bryan95}, we modify the method for use in
hypersonic flows when the thermal energy $e$ is extremely small
compared to the total energy $E$.  This situation presents a problem
because in the total energy method the temperature is computed by
subtracting one large number from another (i.e. the kinetic energy
from the total energy), which tends to generate large numerical
inaccuracies. We overcome this difficulty by additionally solving a
thermal energy equation and using $e$ from this equation when we
expect the error to be large.

@@

\paragraph{Godunov MHD with Constrained Transport (MHD)}
This MHD method is second order in time and space, and preserves the
divergence constraint, $\div \vecB = 0$, to machine precision
through the Constrained Transport (CT) method.  CT, originally
described by \citet{Evans88}, updates the magnetic field with the curl
of an electric field, suitably formulated to preserve $\div
\vecB$.  We employ the CT methods described by \citet{Balsara99} and
\citet{Gardiner05} with the second order hyperbolic
solver of \citet{Li08a} and the divergence free AMR scheme of
\citet{Balsara01}.

@@

\enzo\ includes a photon-conserving radiative transfer algorithm that
is based on an adaptive ray-tracing method utilizing the HEALPix
pixelization of a sphere \citep{AbelWandelt02}. Photons are integrated
outwards from sources using an adaptive timestepping scheme that
preserves accuracy in ionization fronts even in the optically-thin
limit. This has been coupled to the chemistry and cooling network to
provide ionization and heating rates on a cell-by-cell basis. The
method is described in detail (including numerous code tests) in
\citet{Wise11_Moray}.

@@

A second option for radiative transfer is a moment-based method that
adds an additional field tracking the radiation energy density.  This
field is evolved using the flux-limited diffusion method, which
transitions smoothly between streaming (optically thin) and opaque
limits and is coupled to a hydrogen-only ionization network.  The
resulting set of linear equations is solved using the parallel HYPRE
framework.  Full details on the \enzo\ implementation of this method
can be found in \citet{ReynoldsHayesPaschosNorman2009}.

@@

A family of simple heuristic methods are used to model the formation
of stars and their feedback of metals and energy into the gas.  These
methods are based on the work of \citet{CO1992}, and involve the
identification of plausible sites of star formation based on a set of
criteria (for example, dense gas with a short cooling time, which is
both collapsing and unstable).  The local star formation rate is
computed using a range of methods, such as a density-dependent method
based on the Schmidt-Kennicutt relation \citep{K89}.  The affected gas
is converted into a star particle over a few dynamical times, and
metals and thermal energy are injected into the region surrounding the
star particle.

>>>>

In addition, \enzo\ can solve the mass conservation equations for a set of
chemical species and their reactions.  For any species $i$ with comoving
number density $n_i$, these equations have the form:
\begin{equation}
  \frac{\partial n_i}{\partial t} 
  + \frac{1}{a} \div (n_i \vecv) = 
  \sum_j k_{ij}(T) \, n_i n_j + \sum_j \Gamma^{\rm ph}_j n_j 
  \label{eq:species_evolution}
\end{equation}
where $k_{ij}$ are the rate coefficients for the two-body reactions
and are usually functions of only the gas species (we will
specifically note the cases where we either include three-body
reactions or have density-dependent rates).  The $\Gamma^{\rm ph}_j$
are destruction/creation rates due to photoionizations and/or
photodissociations.  Currently the species \enzo\ can follow include
H, H$^+$, He, He$^+$, He$^{++}$, and e$^-$, and through additional
options also H$^-$, H$_2$, H$_2^+$, and HD, D, and D$^+$. Lastly,
\enzo\ can also track the advection of one or more comoving metal density fields,
which can contribute to the radiative cooling and star formation
processes.

@@

In this section we briefly describe the numerical methods which are
used to solve the equations outlined in Section~\ref{sec.equations}.
We proceed through the numerical methods in the same order as will be
used in Sections~\ref{sec.amr} through~\ref{sec.num.analysis} so that
there is a one-to-one correspondence between each of the following
overviews and the complete description provided in the later sections.
The goal in this section is to introduce the reader to the basic
principles of the methods without drowning them in detail.

@@

As described in \citet{1995CoPhC..89..149B}, we modify the method for use in
hypersonic flows when the thermal energy $e$ is extremely small
compared to the total energy $E$.  This situation presents a problem
because in the total energy method the temperature is computed by
subtracting one large number from another (i.e. the kinetic energy
from the total energy), which tends to generate large numerical
inaccuracies. We overcome this difficulty by additionally solving a
thermal energy equation and using $e$ from this equation when we
expect the error to be large.

@@

\paragraph{Godunov MHD with Constrained Transport (MHD)}
This MHD method is second order in time and space, and preserves the
divergence constraint, $\div \vecB = 0$, to machine precision
through the Constrained Transport (CT) method \citep{Collins10}.  CT, originally
described by \citet{Evans88}, updates the magnetic field with the curl
of an electric field, suitably formulated to preserve $\div
\vecB$.  We employ the CT methods described by \citet{Balsara99} and
\citet{Gardiner05} with the second order hyperbolic
solver of \citet{Li08a} and the divergence free AMR scheme of
\citet{Balsara01}.

@@

\enzo\ includes a photon-conserving radiative transfer algorithm that
is based on an adaptive ray-tracing method utilizing the HEALPix
pixelization of a sphere \citep{Abel02_RT}. Photons are integrated
outwards from sources using an adaptive timestepping scheme that
preserves accuracy in ionization fronts even in the optically-thin
limit. This has been coupled to the chemistry and cooling network to
provide ionization and heating rates on a cell-by-cell basis. The
method is described in detail (including numerous code tests) in
\citet{Wise11_Moray}.

@@

A second option for radiative transfer is a moment-based method that
adds an additional field tracking the radiation energy density.  This
field is evolved using the flux-limited diffusion method, which
transitions smoothly between streaming (optically thin) and opaque
limits and is coupled to an ionization network of either purely
hydrogen, or both hydrogen and helium.  The
resulting set of linear equations is solved using the parallel HYPRE
framework.  Full details on the \enzo\ implementation of this method
can be found in \citet{ReynoldsHayesPaschosNorman2009}.

@@

A family of simple heuristic methods are used to model the formation
of stars and their feedback of metals and energy into the gas.  These
methods are based on the work of \citet{CO1992}, and involve the
identification of plausible sites of star formation based on a set of
criteria (for example, dense gas with a short cooling time, which is
both collapsing and unstable).  The local star formation rate is
computed using a range of methods, such as a density-dependent method
based on the Schmidt-Kennicutt relation \citep{K89}.  The affected gas
is converted into a star particle over a few dynamical times, and
metals and thermal energy are injected into the region surrounding the
star particle.  A related set of methods involves the simulation of
single Population III stars, rather than ensembles, and is calibrated
by \textit{ab initio} simulations of primordial star formation.

### 2013-06-20 table_of_variables_and_constants.tex
Mike Norman's edits.

  \multicolumn{2}{l}{\underline{Natural Constants and Values}} \\[5pt]
  $G$ & gravitational constant \\
  $H$ & Hubble constant \\
  $k_B$ & Boltzmann constant \\
  $M_\odot$ & mass of Sun \\
  $m_H$ & mass of hydrogen atom \\
  $m_e$ & mass of electron \\
  $c$ & speed of light \\
  $c_{\rm s}$ & sound speed \\
  \enddata

>>>>

  \multicolumn{2}{l}{\underline{Natural Constants and Values}} \\[5pt]
  $G$ & gravitational constant \\
  $H$ & Hubble constant \\
  $k_B$ & Boltzmann constant \\
  $M_\odot$ & mass of Sun \\
  $m_H$ & mass of hydrogen atom \\
  $m_e$ & mass of electron \\
  $c$ & speed of light \\
  $c_{\rm s}$ & sound speed \\
  $z$ & redshift \\
  \enddata

### 2013-06-20 tests-zeldovichpancake.tex
Mike Norman's edits.

The final output of the calculation, with the key features of this
test problem, is shown in Figure~\ref{fig.pancake}.
The strong shocks and large density gradients are well-resolved, with
density and velocity jumps being well-delineated and at the correct
locations.  The key features of 
this test problem can be resolved with far fewer cells -- simulations
including a mere 8 cells resolve the key features, as shown in
Sections 3.3.4-3.3.5 of~\citet{1996PhDT........80B} -- but we choose a
higher resolution here for illustrative purposes.

>>>>

The final output of the calculation, with the key features of this
test problem, is shown in Figure~\ref{fig.pancake}.
The strong shocks and large density gradients are well-resolved, with
density and velocity jumps being well-delineated and at the correct
locations.  The key features of 
this test problem can be resolved with far fewer cells -- simulations
including a mere 8 cells resolve the key features, as shown in
Sections 3.3.4-3.3.5 of~\citet{BryanThesis96} -- but we choose a
higher resolution here for illustrative purposes.
